---
title: Portfolio Possibilities Curve (PPC)
description: PPC with GPT4 as my coding partner
author: David Harper, CFA, FRM
date: 2023-11-19
categories: [code, analysis]
execute: 
  echo: true
  warning: false
---

Contents

* Define the sets of stocks/indices
* Retrieve returns for each periodicity
* Analyze: add the analysis list column
* Setup the simulation
* Run the simulation
* Visualize

load libraries

```{r}

library(tidyverse)
library(tidyquant)
# library(dplyr); library(tidyr); library(purrr)
# library(ggplot)

```


## Define the SETS of stocks/indices

The container in this approach is **stock_sets**, a dataframe the we initialize (our TOC) with three columns:

* set_id
* description
* symbols: a list of tickers

```{r}

sector_3eft_list <- c( "XLK", "XLV", "XLP") # Tech, Health, Staples
# sector_4etf_list <- c( "XLK", "XLV", "XLP",  "XLE",) # Tech, Health, Staples, Energy
sector_5etf_list <- c( "XLK", "XLV", "XLP", "XLE", "XLF") # Tech, Health, Staples, Energy, Financials
# sector_7etf_list <- c(sector_5etf_list, "XLI", "XLU")

size_etfs <- c("SPY", "MDY", "IWM") # Large, Mid, Small
# size_style_etfs <- c("IWF",  # Large-Cap Growth
#                      "IWD",  # Large-Cap Value
#                      "SPY",  # Large-Cap Blend
#                      "IWP",  # Mid-Cap Growth
#                      "IWS",  # Mid-Cap Value
#                      "MDY",  # Mid-Cap Blend
#                      "IWO",  # Small-Cap Growth
#                      "IWN",  # Small-Cap Value
#                      "IWM")  # Small-Cap Blend

stock_sets <- tibble(
    set_id =      c("3_Sectors", "5_Sectors","3_Sizes"),
    description = c("3 Sectors picked by GPT-4: Tech, Health, Staples",
                    "5 Sectors picked by GPT-4: above + Energy + Financials",
                    "Size: Large, Mid, Small--Blend"),
    
    # this is a list column, see https://adv-r.hadley.nz/vectors-chap.html#list-columns 
    symbols = list(sector_3eft_list, sector_5etf_list, size_etfs)
    )

# c(                    "XLK",  # Technology
#                       "XLV",  # Health Care
#                       "XLF",  # Financials
#                       "XLY",  # Consumer Discretionary
#                       "XLP",  # Consumer Staples
#                       "XLE",  # Energy
#                       "XLU",  # Utilities
#                       "XLI",  # Industrials
#                       "XLB",  # Materials
#                       "XLRE", # Real Estate
#                       "XLC") # Communication Services

date_start <- "2013-01-01"
date_end   <- "2023-11-17"

```

## Retrieve returns for each periodicity; aka, frequency

For each SET of tickers, **get_returns** will retrieve log returns for each of three periods:

* daily
* weekly
* monthly

Then we will call the get_returns function via map (my favorite function) to create a new list column called **nested_data**. Each row of nested_data *will contain a list* of three dataframes, one for each period. These dataframes will contain the log returns for each ticker in the set. 

```{r}

get_returns <- function(symbols, start_date, end_date) {
    mult_stocks <- tq_get(symbols, get = "stock.prices", 
                          from = start_date, to = end_date)
    
    periods <- c("daily", "weekly", "monthly")
    returns_list <- lapply(periods, function(period) {
        mult_stocks |> 
            group_by(symbol) |> 
            tq_transmute(select = adjusted,
                         mutate_fun = periodReturn, 
                         period = period, 
                         type = "log")
    })
    
    names(returns_list) <- periods
    return(returns_list)
}

# Nest return data for each stock set
stock_sets <- stock_sets |> 
    mutate(nested_data = map(symbols, 
                             ~ get_returns(.x, date_start, date_end)))

print(stock_sets)

```

## Analyze: add the analysis list column

For each set and periodicity, the analysis list column generates:

* vector of volatilities
* vector of average returns
* correlation matrix (diagonal is 1)
* average correlation (as a *rough* measure of diversification)


```{r}

perform_analysis <- function(data, returns_column) {
    volatilities <- data |>  
        group_by(symbol) |>  
        summarise(volatility = sd(.data[[returns_column]], na.rm = TRUE)) |>  
        ungroup()
    
    avg_returns <- data |>  
        group_by(symbol) |>  
        summarise(avg_return = mean(.data[[returns_column]], na.rm = TRUE)) |>  
        ungroup()
    
    data_wide <- data |>  
        pivot_wider(names_from = symbol, values_from = .data[[returns_column]])
    corr_matrix <- cor(select(data_wide, -date), use = "complete.obs")
    avg_corr <- mean(corr_matrix[lower.tri(corr_matrix)])
    
    return(list(volatilities = volatilities, avg_returns = avg_returns, corr_matrix = corr_matrix, avg_corr = avg_corr))
}

# Applying the perform_analysis function to the stock_sets
stock_sets <- stock_sets |> 
    mutate(analysis = map(nested_data, ~ {
        data_daily <- .x$daily
        data_weekly <- .x$weekly
        data_monthly <- .x$monthly
        
        analysis_daily <- perform_analysis(data_daily, "daily.returns")
        analysis_weekly <- perform_analysis(data_weekly, "weekly.returns")
        analysis_monthly <- perform_analysis(data_monthly, "monthly.returns")
        
        list(daily = analysis_daily, weekly = analysis_weekly, monthly = analysis_monthly)
    }))

# Examine data structure 
print(stock_sets) # Notice the analysis list column has been added
glimpse(stock_sets)
stock_sets$analysis[[1]] # first row is the first stock set

```

Here is a class diagram of the stock_sets data structure rendered by [DiagrammeR via Graphviz](https://graphviz.org/Gallery/directed/UML_Class_diagram.html).

```{r echo=FALSE}

library(DiagrammeR)

graphviz_script <- "
digraph stock_sets_structure {
    graph [splines=ortho]
    
    node [shape=record]
    
    stock_sets [label=\"{stock_sets|+ set_id : String\\l+ description : String\\l+ symbols : List\\l+ nested_data : List\\l+ analysis : List\\l}\"]

    nested_data [label=\"{nested_data|+ daily : DataFrame\\l+ weekly : DataFrame\\l+ monthly : DataFrame\\l}\"]

    analysis [label=\"{analysis|+ daily_analysis : List\\l+ weekly_analysis : List\\l+ monthly_analysis : List\\l}\"]

    daily_analysis [label=\"{daily_analysis|+ volatilities : DataFrame\\l+ avg_returns : DataFrame\\l+ corr_matrix : Matrix\\l+ avg_corr : Numeric\\l}\"]

    weekly_analysis [label=\"{weekly_analysis|+ volatilities : DataFrame\\l+ avg_returns : DataFrame\\l+ corr_matrix : Matrix\\l+ avg_corr : Numeric\\l}\"]

    monthly_analysis [label=\"{monthly_analysis|+ volatilities : DataFrame\\l+ avg_returns : DataFrame\\l+ corr_matrix : Matrix\\l+ avg_corr : Numeric\\l}\"]

    stock_sets -> nested_data
    stock_sets -> analysis
    analysis -> daily_analysis
    analysis -> weekly_analysis
    analysis -> monthly_analysis
}
"

DiagrammeR(graphviz_script, type = "graphviz")

```

## Setup the simulation

The **get_random_weights** function returns a dataframe of random weights. Each column is a set of weights for a single simulation. Each row is the weight for a single stock. The weights are normalized so that they sum to 1. 

So I'm starting with an incredibly naive approach to the simulation. I'm going to assume that the expected return for each stock is the average return for that stock over the entire period. I'm also going to assume that the volatility for each stock is the average volatility for that stock over the entire period. *Most importantly*, the only randomness in the simulation is the weights and they are totally naive because they are independent of the analysis. We can't expect anything like an efficient frontier from the raw scatterplot. However, this will illustrate the future risk/reward trade-off faced by a "totally naive" investor!

```{r}

# returns a data frame of random weights
# rows = weight per stock; columns = number of simulations
get_random_weights <- function(num_stocks, num_simulations) {
    set.seed(123)
    weights_df <- matrix(nrow = num_stocks, ncol = num_simulations)

    for (i in 1:num_simulations) {
        weights <- runif(num_stocks)
        weights_df[, i] <- weights / sum(weights)
    }

    return(as.data.frame(weights_df))
}

# single simulation: given a set of weights, computes the expected return and volatility
port_sim <- function(exp_returns, volatilities, corr_matrix, weights) {
    
    cov_matrix <- outer(volatilities, volatilities) * corr_matrix
    port_variance <- t(weights) %*% cov_matrix %*% weights
    port_exp_return <- sum(weights * exp_returns)

    return(list(exp_returns = exp_returns, 
                volatilities = volatilities,
                cov_matrix = cov_matrix, 
                corr_matrix = corr_matrix,
                port_variance = port_variance,
                port_exp_return = port_exp_return))
}

# runs a port_simulation for each column in the weights_df
run_sims <- function(exp_returns, volatilities, corr_matrix, weights_df) {
    simulations <- map(1:ncol(weights_df), ~ {
        weights_vector <- weights_df[, .x]
        port_sim(exp_returns, volatilities, corr_matrix, weights_vector)
        })
    
    return(simulations)
}

```

## Run the simulation (on a single set)

```{r}

# Selecting the desired set (e.g., "Set 1")
select_set <- stock_sets |> 
    filter(set_id == "5_Sectors") |> 
    pull("analysis")

analyze_set <- select_set[[1]]
analyze_period <- analyze_set$monthly

# Extracting components from the selected set
exp_returns_period <- analyze_period$avg_returns$avg_return
volatilities_period <- analyze_period$volatilities$volatility
corr_matrix_period <- analyze_period$corr_matrix
num_stocks_period <- length(volatilities_period)

num_sims <- 20000  # Set the number of simulations
random_weights_df_period <- get_random_weights(num_stocks_period, num_sims)
sim_results_period <- run_sims(exp_returns_period, 
                              volatilities_period, 
                              corr_matrix_period, 
                              random_weights_df_period)

# Print results of the first simulation
print(sim_results_period[[1]])

results_df_period <- map_dfr(sim_results_period, ~ data.frame(Exp_Return = .x$port_exp_return, 
                                                            Std_Dev = sqrt(.x$port_variance)))
# View summarized results for daily returns
print(head(results_df_period))
results_df <- results_df_period

```

## Visualize the results

```{r}

library(patchwork)

results_df <- results_df |> 
    arrange(Std_Dev) |> 
    mutate(is_efficient = Exp_Return >= cummax(Exp_Return))

efficient_portfolios <- results_df |> 
    arrange(Std_Dev)  |> 
    mutate(cummax_return = cummax(Exp_Return)) |> 
    filter(Exp_Return >= cummax_return)

efficient_model <- lm(Exp_Return ~ poly(Std_Dev, 2), data = efficient_portfolios)

p1 <- ggplot(results_df, aes(x = Std_Dev, y = Exp_Return, color = is_efficient)) +
    geom_point() +
    scale_color_manual(values = c("azure2", "springgreen4")) + 
    theme_minimal() +
    theme(
        axis.title = element_blank(),
        legend.position = "none"
    )

p2 <- ggplot(results_df, aes(x = Std_Dev, y = Exp_Return)) +
    geom_point(aes(color = is_efficient), size = 1) +  # Default size for all points
    geom_point(data = filter(results_df, is_efficient), 
               aes(color = is_efficient), size = 2) +  # Larger size for efficient points
    scale_color_manual(values = c("azure2", "springgreen4")) +
    theme_minimal() +
    geom_line(data = efficient_portfolios, aes(x = Std_Dev, y = Exp_Return), colour = "springgreen2") +
    theme(
        axis.title = element_blank(),
        legend.position = "none"
    )

p3 <- ggplot(results_df, aes(x = Std_Dev, y = Exp_Return)) +
    geom_point(color = "azure2") +
    geom_smooth(data = efficient_portfolios, method = "lm", formula = y ~ poly(x, 2), 
                se = FALSE, colour = "springgreen4", linewidth = 1.5) +
    labs(x = "Std Dev (Risk)",
         y = "Return") +
    theme_minimal()

# Calculate a color metric based on Exp_Return and Std_Dev

RiskFree_temp <- 0.0
results_df <- results_df %>%
    mutate(efficiency = (Exp_Return - RiskFree_temp)/ Std_Dev)

# Create a scatterplot with color gradient based on the color_metric
# p4 <- ggplot(results_df, aes(x = Std_Dev, y = Exp_Return, color = color_metric)) +
#     geom_point() +
#     scale_color_gradient2(low = "azure3", high = "springgreen1", mid = "yellow", 
#                           midpoint = median(results_df$color_metric)) +
#     theme_minimal() +
#     labs(color = "Color Metric")

# Assuming results_df and color_metric are already defined appropriately

p4 <- ggplot(results_df, aes(x = Std_Dev, y = Exp_Return, color = efficiency)) +
    geom_point() +
    scale_color_gradientn(colors = c("azure4", "lightgoldenrod1", "springgreen2"),
                          values = scales::rescale(c(min(results_df$efficiency), 
                                                     max(results_df$efficiency)))) +
    theme_minimal() +
    labs(x = "Std Dev (Risk)",
         y = "Return", 
         color = "efficiency") 

(p1 + p2) / (p3 + p4 )

```

